---
title: "Homework 1: Environmental Injustice"
execute: 
  warning: false
  message: false
format: html
---

```{r}
# Load packages
pacman::p_load('tidyverse', 'sf', 'here', 'tmap', 'tmaptools')
```

```{r}
# Load data with sf package
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))
```

# Georgia, USA

```{r}
# filter to georgia state
georgia <- ejscreen |> 
  filter(ST_ABBREV == "GA") 

# filter to Bacon, Georgia
bacon <- georgia |> 
  filter(CNTY_NAME %in% c("Bacon County"))

# find average values for all variables within counties
georgia_counties_means <- aggregate(georgia, by = list(georgia$CNTY_NAME), FUN = mean)
```

```{r}
# Define extent with st_bbox()
bbox_georgia <- st_bbox(c(xmin = -85.9, xmax = -80.5, ymax = 35.1, ymin = 30.4), crs = 4326)
```

```{r}
# Create first map
tm_shape(georgia_counties_means, bbox = bbox_georgia) +
  tm_crs(4326) + # Specify crs
  tm_polygons(fill = "PEOPCOLORPCT",
              fill.scale = tm_scale(values = 'brewer.yl_or_rd'),
              fill.legend = tm_legend(title = "Percent People of Color", reverse = TRUE)) +
  tm_title(text = "Georgia Counties: Percent People of Color", fontface = 'bold', size = 4) +
  tm_compass(position = c(0.76, 0.95), type = '4star') + # Add compass
  tm_scalebar(position = c(0.05, 0.65), text.size = 0.7) + # Add scale bar
  tm_layout(legend.position = c(0, 1), # Adjust legend position
            inner.margins = c(0.05, 0.5, 0.05, 0.05)) 
```

```{r}
# Create second map
tm_shape(georgia_counties_means, bbox = bbox_georgia) +
  tm_crs(4326) +
  tm_polygons(fill = "RESP",
              fill.scale = tm_scale(values = 'brewer.yl_or_rd'),
              fill.legend = tm_legend(title = "Respiratory Hazard Index", reverse = TRUE)) +
  tm_title(text = "Georgia Counties: Respiratory Hazard Index", fontface = 'bold', size = 4) +
  tm_compass(position = c(0.76, 0.95), type = '4star') +
  tm_scalebar(position = c(0.05, 0.5), text.size = 0.7) +
  tm_layout(legend.position = c(0, 1),
            inner.margins = c(0.05, 0.5, 0.05, 0.05))
```

The state of Georgia has some of the nation's worst air quality according to the [American Lung Association](https://www.lung.org/research/sota/city-rankings/states/georgia). Additionally, [Black Demographics](https://blackdemographics.com/population/black-state-population/) has ranked Georgia as second highest percent black population in the US. The first map shows this distribution of percent people of color by county, concentrated around central Georgia and south-western parts as well. The second map shows the distribution of the EPA's AirToxScreen respiratory hazard index, derived from measuring toxic pollutants in ambient outdoor air. The hazard index is a unit-less measurement of the risk for developing non-cancer respiratory illnesses like asthma and bronchitis. The most hazardous counties lie in close proximity to counties with higher percentage of black populations which shows one example of environmental injustice that black communities face in the US. 
